---
title: "MATH630-HW1"
author: "Joshua Burkhart"
date: "October 13, 2015"
output: pdf_document
---
```{r}
library(gapminder)
library(dplyr)
library(ggplot2)
library(grid)
library(tidyr)
library(moments)
library(psych)
library(reshape2)
```

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/", message = FALSE, warning = FALSE, include = TRUE, echo = TRUE, error = TRUE, fig.width = 11, comment = NA)
```

HLO Gapminder

```{r}
str(gapminder)
glimpse(gapminder)
names(gapminder)
head(gapminder)
nrow(gapminder)
ncol(gapminder)
unique(is.na(gapminder))
```
    
    Is it a data.frame, a matrix, a vector, a list?
    data.frame
    
    What is the unit of analysis in the dataset?
    Excerpt of the Gapminder data on life expectancy, GDP per capita, and population by country, every five years, from 1952 to 2007
    from http://www.gapminder.org/data/
    
    How many variables/columns?
    6
    
    How many rows/observations?
    1704
    
    Which variables are continuous?
    "lifeExp" "gdpPercap"
    
    Which variables are discrete?
    "country" "continent" "pop" "year"
    
    Which variables are categorical?
    "country" "continent"
    
    How many levels do they have?
    country: 142
    continent: 5
    
    What about missing data for any variables?
    no missing data reported

Numerical and counting detective work

```{r}
summary(gapminder)
ggplot(gapminder,aes(lifeExp)) + geom_histogram(color = "white")
```

    Pick one quantitative variable to explore using descriptive statistics as discussed in class.
    lifeExp
    
        Characterize the range of possible values, max vs. min, etc.- does it make sense?
        Min: 23.6
        1st Q: 48.2
        Median: 60.71
        Mean: 59.47
        3rd Q: 70.85
        Max: 82.6
        
        These values make sense.
        
        What's the center? What's the spread? What's the shape?
        Feel free to use summary statistics or tables. You don't need to re-summarise summarised data for us. It is one thing to be able to get R to give you what you ask for. It is another to interpret what R gives you. We are more interested in the latter here, but also that you can do the former without errors.
        graph looks bimodal, higher peak to the right, skews down
        
        Comment on representativeness of measures of central tendency, given the spread and shape.
        IQR / median / mean don't hint at the bimodal distribution but
        mean < median does hint at skew down
        
    Pick one categorical variable and generate the n's (in whatever the appropriate "unit of analysis" is) and proportions of the sample that contribute to each level of that variable.
    continent
    
    Africa  :624 = 0.3661972
    Americas:300 = 0.1760563
    Asia    :396 = 0.2323944
    Europe  :360 = 0.2112676
    Oceania : 24 = 0.01408451
    
```{r}
filter(gapminder, continent=="Africa") %>% nrow / nrow(gapminder)
filter(gapminder, continent=="Americas") %>% nrow / nrow(gapminder)
filter(gapminder, continent=="Asia") %>% nrow / nrow(gapminder)
filter(gapminder, continent=="Europe") %>% nrow / nrow(gapminder)
filter(gapminder, continent=="Oceania") %>% nrow / nrow(gapminder)
```

    Which level contains the smallest number of observations? The largest?
    smallest: Oceana
    largest: Africa
    
    Generate your descriptive statistics again, now stratified by the different levels of your categorical variable.
    
```{r}
africaPlot <- gapminder %>% filter(continent=="Africa") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Africa")

americasPlot <- gapminder %>% filter(continent=="Americas") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Americas")

asiaPlot <- gapminder %>% filter(continent=="Asia") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Asia")

europePlot <- gapminder %>% filter(continent=="Europe") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Europe")

oceaniaPlot <- gapminder %>% filter(continent=="Oceania") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Oceania")

#from http://stackoverflow.com/questions/24387376/r-weird-error-could-not-find-function-multiplot
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

multiplot(africaPlot,americasPlot,asiaPlot,europePlot,oceaniaPlot, cols=3)
gapminder %>% filter(continent=="Africa") %>% summary
gapminder %>% filter(continent=="Americas") %>% summary
gapminder %>% filter(continent=="Asia") %>% summary
gapminder %>% filter(continent=="Europe") %>% summary
gapminder %>% filter(continent=="Oceania") %>% summary

```

    
        How did any of your initial observations of the quantitative variable change?
        Foreshadowing: look for differences in both center and spread across categories. Think about what this means in terms of possible comparisons between means across different levels of that factor.
        Asia, Americas, and Europe look similar in that they all skew down.
        Africa and Oceania skew up. The plots indicate the means of Asia, 
        Americas, and Europe may be lower than their medians, while the
        plots of Africa and Oceania indicate their means may be higher
        than their medians.

Graphical detective work

    Graphically explore your one quantitative variable using histograms and boxplots. See the exploratory data analysis link for example R code.

```{r}
africaPlot <- gapminder %>% filter(continent=="Africa") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Africa")
baf <- gapminder %>% filter(continent=="Africa") %>% ggplot(aes(x="Africa",y=lifeExp)) + geom_boxplot()
americasPlot <- gapminder %>% filter(continent=="Americas") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Americas")
bam <- gapminder %>% filter(continent=="Americas") %>% ggplot(aes(x="Americas",y=lifeExp)) + geom_boxplot()
asiaPlot <- gapminder %>% filter(continent=="Asia") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Asia")
bas <- gapminder %>% filter(continent=="Asia") %>% ggplot(aes(x="Asia",y=lifeExp)) + geom_boxplot()
europePlot <- gapminder %>% filter(continent=="Europe") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Europe")
beu <- gapminder %>% filter(continent=="Europe") %>% ggplot(aes(x="Europe",y=lifeExp)) + geom_boxplot()
boc <- gapminder %>% filter(continent=="Oceania") %>% ggplot(aes(x="Oceania",y=lifeExp)) + geom_boxplot()
oceaniaPlot <- gapminder %>% filter(continent=="Oceania") %>% ggplot(aes(lifeExp)) + geom_histogram(color = "white") + ggtitle("Oceania")

multiplot(baf,africaPlot,
          bam,americasPlot,
          bas,asiaPlot,
          beu,europePlot,
          boc,oceaniaPlot,
          cols=5)
```


        What are you looking for in each plot?
        I'm looking for the IQR (box height) and median (where the box is centered).
        
        Do you notice anything interesting/puzzling/surprising?
        The box plots seem to match the histograms (duh) but Europe's box plot looks like the odd one out as it has such a small Q2-Q3 range and such a high median. Also, Oceania's Q2-Q3 looks unusually large.

>>>

        Look back at your descriptive statistics for your variable. Comment on the descriptive value of the numbers in light of your visualizations. 
        The numbers aren't wrong, they just don't make the differences as apparent.
        
        Do a quick sanity check- does everything look consistent across numerical and graphical depictions of your data?
        yes
        
    Add your one categorical variable to the mix and graphically explore your quantitative variable using any of the combination plots discussed in class. Your new plot must account for the categorical variable, either by facetting by levels of that variable, setting an aesthetic (color, shape, etc.) to differ across levels, or stratifying the x-axis by the different levels of your categorical variable. See the exploratory data analysis link for ideas. We want to see you exploring multiple types of plots, and each plot should include at least 2 "layers" of information. Sampling 100 random rows from the dataset is a valid strategy here (reference last slide from Class 2 EDA class) if you want to compare big n/small n types of plots.
    
```{r}

vaf <- gapminder %>% sample_n(20) %>% filter(continent=="Africa") %>% ggplot(aes(x="Africa",y=lifeExp)) + geom_violin()+ geom_boxplot(width=0.1)+coord_cartesian(ylim = c(25,85))
vam <- gapminder %>% sample_n(20) %>% filter(continent=="Americas") %>% ggplot(aes(x="Americas",y=lifeExp)) + geom_violin()+ geom_boxplot(width=0.1)+coord_cartesian(ylim = c(25,85))
vas <- gapminder %>% sample_n(20) %>% filter(continent=="Asia") %>% ggplot(aes(x="Asia",y=lifeExp)) + geom_violin()+ geom_boxplot(width=0.1)+coord_cartesian(ylim = c(25,85))
veu <- gapminder %>% sample_n(20) %>% filter(continent=="Europe") %>% ggplot(aes(x="Europe",y=lifeExp)) + geom_violin()+ geom_boxplot(width=0.1)+coord_cartesian(ylim = c(25,85))
voc <- gapminder %>% filter(continent=="Oceania") %>% ggplot(aes(x="Oceania",y=lifeExp)) + geom_violin()+ geom_boxplot(width=0.1)+coord_cartesian(ylim = c(25,85))

multiplot(vaf,
          vam,
          vas,
          veu,
          voc,
          cols=5)

```

```{r}
gapminder %>% ggplot(aes(colour=factor(continent),x=year,y=lifeExp)) + geom_smooth()
```

In-depth detective work

Manipulate and further explore the gapminder dataset with the dplyr package, complemented by visualizations made with ggplot2. Pick at least two of the tasks below from the task menu and approach each with a table and figure.

    dplyr should be your main data manipulation tool
    ggplot2 should be your main visualization tool

Make observations about what your tables/figures show and about the process. If you want to do something comparable but different, i.e. swap one quantitative variable for another- go for it!

You do not have to use tidyr or otherwise worry about reshaping your tables. Many of your tables may not be formatted perfectly in the report. Simply printing dplyr tabular output is fine. For all things, graphical and tabular, if you're dissatisfied with a result, discuss the problem, what you tried to do to fix it, and move on.

Task menu

    Get the maximum and minimum of GDP per capita for all continents.

```{r}
max_min_gdp <- gapminder %>% group_by(continent) %>% summarize(min(gdpPercap), max(gdpPercap))
max_min_gdp_stack <- max_min_gdp %>% melt(id="continent")
max_min_gdp_stack %>% ggplot(aes(x=continent, y=value, fill=variable)) +
  geom_bar(stat="identity") +
  geom_text(aes(label = round(value)), size = 7) + theme(legend.position="bottom")
max_min_gdp
```

    Look at the spread of GDP per capita across countries within the continents.

```{r}
max_min_sum <- gapminder %>% filter(continent=="Africa") %>% group_by(country) %>% summarize(min(gdpPercap), max(gdpPercap))
max_min_sum
country_stack <- max_min_sum %>% melt(id=c("country"))
ordered_stack <- country_stack
ordered_stack$country <- factor(country_stack$country,levels=country_stack[order(country_stack$value),"country"])
oaf <- ordered_stack %>% ggplot(aes(x=country, y=value, fill=variable)) + geom_bar(stat="identity") + coord_flip() + theme(legend.position="top")

max_min_sum <- gapminder %>% filter(continent=="Asia") %>% group_by(country) %>% summarize(min(gdpPercap), max(gdpPercap))
max_min_sum
country_stack <- max_min_sum %>% melt(id=c("country"))
ordered_stack <- country_stack
ordered_stack$country <- factor(country_stack$country,levels=country_stack[order(country_stack$value),"country"])
oas <- ordered_stack %>% ggplot(aes(x=country, y=value, fill=variable)) + geom_bar(stat="identity") + coord_flip() + theme(legend.position="top")

max_min_sum <- gapminder %>% filter(continent=="Americas") %>% group_by(country) %>% summarize(min(gdpPercap), max(gdpPercap))
max_min_sum
country_stack <- max_min_sum %>% melt(id=c("country"))
ordered_stack <- country_stack
ordered_stack$country <- factor(country_stack$country,levels=country_stack[order(country_stack$value),"country"])
oam <- ordered_stack %>% ggplot(aes(x=country, y=value, fill=variable)) + geom_bar(stat="identity") + coord_flip() + theme(legend.position="top")

max_min_sum <- gapminder %>% filter(continent=="Europe") %>% group_by(country) %>% summarize(min(gdpPercap), max(gdpPercap))
max_min_sum
country_stack <- max_min_sum %>% melt(id=c("country"))
ordered_stack <- country_stack
ordered_stack$country <- factor(country_stack$country,levels=country_stack[order(country_stack$value),"country"])
oeu <- ordered_stack %>% ggplot(aes(x=country, y=value, fill=variable)) + geom_bar(stat="identity") + coord_flip() + theme(legend.position="top")

max_min_sum <- gapminder %>% filter(continent=="Oceania") %>% group_by(country) %>% summarize(min(gdpPercap), max(gdpPercap))
max_min_sum
country_stack <- max_min_sum %>% melt(id=c("country"))
ordered_stack <- country_stack
ordered_stack$country <- factor(country_stack$country,levels=country_stack[order(country_stack$value),"country"])
ooc <- ordered_stack %>% ggplot(aes(x=country, y=value, fill=variable)) + geom_bar(stat="identity") + coord_flip() + theme(legend.position="top")

multiplot(oaf,
          oam,
          oas,
          oeu,
          ooc,
          cols=5)

```


    How does life expectancy vary across different continents?
    
```{r}
max_min_sum <- gapminder %>% group_by(continent) %>% summarize(min(lifeExp), max(lifeExp))
continent_stack <- max_min_sum %>% melt(id="continent")
continent_stack %>% ggplot(aes(x=continent, y=value, fill=variable)) +
  geom_bar(stat="identity") +
  geom_text(aes(label = round(value)), size = 7)
max_min_sum
```

    Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy - you decide - a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

```{r}
#group by year
g_by_year <- gapminder %>% melt(id=c("year","lifeExp","country"))

#low lifeExp = lifeExp < median for year
median_by_year <- g_by_year %>% group_by(year) %>% summarize(median(lifeExp))

#num countries where lifeExp < low lifeExp
num_c_by_year <- g_by_year %>% group_by(year) %>% summarise(count = length(country[lifeExp < median_by_year$`median(lifeExp)`]))

#x=year,y=num_countries
num_c_by_year %>% ggplot(aes(fill="red",x=year,y=count)) + geom_bar(stat="identity",width=3) + ggtitle("Number of countries below median lifeExp for the year")

num_c_by_year
```


    Make up your own! Look back at our Class 2 slides for dplyr example ideas with the diamonds dataset, and the package vignettes for other ideas.

Companion graphs

For each table, make sure to include a relevant figure. One tip for starting is to draw out on paper what you want your x- and y-axis to be first and what your geom is; that is, start by drawing the plot you want ggplot to give you. Your figure does not have to depict every single number present in the table. Use your judgement. It just needs to complement the table, add context, and allow for some sanity checking.

further examining distribution of lifeExp accross continents
```{r}
max_min_sum <- gapminder %>% group_by(continent) %>% summarize(LE_skewness=moments::skewness(lifeExp))
continent_stack <- max_min_sum %>% melt(id="continent")
skew <- continent_stack %>% ggplot(aes(x=continent, y=value, fill=variable)) +
  geom_bar(stat="identity") +
  geom_text(aes(label = round(value, digits=2)), size = 7) + ggtitle("Life Expectancy Skewness by Continent") + theme(legend.position="bottom")
max_min_sum

max_min_sum <- gapminder %>% group_by(continent) %>% summarize(LE_kurtosis=moments::kurtosis(lifeExp))
continent_stack <- max_min_sum %>% melt(id="continent")
kurt <- continent_stack %>% ggplot(aes(x=continent, y=value, fill=variable)) +
  geom_bar(stat="identity") +
  geom_text(aes(label = round(value, digits=2)), size = 7) + ggtitle("Life Expectancy Kurtosis by Continent") + theme(legend.position="bottom")
max_min_sum

multiplot(skew,kurt,cols=2)
```


Notice which figures are easy/hard to make, and whether the visualization adds clarity, detracts from, or is completely redundant (and therefore probably unnecessary) with respect to the tabular display.

The two most time-consuming plots / tables to generate involved 
1. ordering bars by a value other than their label (ordering countries by min life exp)
2. grouping items into a variable by condition (counting countries w < median life exp)

Report your process

1. consider how a table or plot should look
2. review documentation on https://rpubs.com/bradleyboehmke/data_wrangling
3. guess at what to do based on prior knowledge
4. google frantically (typically landing somewhere on stackoverflow.com or rpubs.com)
5. apply lessons learned from forums to my current problem
6. wrangle my data into a sufficient-looking table
7. repeat steps 3-5 to produce plot

You're encouraged to reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc.
