---
title: "midterm"
author: "Joshua Burkhart"
output: 
  pdf_document: 
    latex_engine: xelatex
---
```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
```

Midterm: Simple Linear Regression
=================================

Overview
--------
```{r}
library(plyr) #this must be loaded before dplyr 
library(dplyr)
library(ggplot2)
library(broom)
```

```{r}
library(Lahman)
?Lahman
```

```{r}
Teams <- Lahman::Teams
Salaries <- Lahman::Salaries
```

HLO Lahman
----------

```{r}
str(Teams)
str(Salaries)
glimpse(Teams)
glimpse(Salaries)
head(Teams)
head(Salaries)
tail(Teams)
tail(Salaries)
names(Teams)
names(Salaries)
ncol(Teams)
ncol(Salaries)
length(Teams)
length(Salaries)
head(rownames(Teams))
head(rownames(Salaries))
dim(Teams)
dim(Salaries)
nrow(Teams)
nrow(Salaries)
summary(Teams)
summary(Salaries)
?Teams
?Salaries
```

Are they data.frames, matrices, vectors, lists?

> data.frames  

What is the unit of analysis in the dataset?

> Teams: Yearly statistics and standings for teams.  
  Salaries: Player salary data.  

How many variables/columns?

> Teams: 48  
  Salaries: 5  

How many rows/observations?

> Teams: 2775  
  Salaries: 24758  

Find the variables for games won, team, year, and salary.

> W: Wins  
  teamID: Team; a factor  
  yearID: Year  
  salary: Salary  

Which variables are continuous?

> In theory, salary could be continuous but in practice, salary looks like it's rounded to the nearest thousand.  

Which variables are discrete?

> W  
  teamID  
  yearID  
  salary  

Which variables are categorical?

> teamID  

How many levels do they have?

> 149  

What about missing data for any variables?

```{r}
unique(is.na(Teams))
unique(is.na(Salaries))
```

> Teams: Missing data in several columns  
  Salaries: No missing data reported  
  
Data wrangling in dplyr
-----------------------

Univariate EDA (+ more wrangling)
---------------------------------

Bivariate EDA (+ even more wrangling)
-------------------------------------

Regression model
----------------

Wrap-up
-------


  
