---
title: "non_parametric"
author: "Joshua Burkhart"
date: "November 24, 2015"
output: 
  pdf_document: 
    latex_engine: xelatex
---

# Class 16: Nonparametric methods for two sample problems

```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)
library(car)
library(datasets)
library(coin)
library(pairwiseCI)
options(scipen=999)
```

## Parametric tests of homogeneity of variances

```{r}
set.seed(10)
low <- rnorm(30, 0, 3) 
med <- rnorm(30, 0, 6)
high <- rnorm(30, 0, 12)
var_df <- data.frame(low, med, high)
var_df_plot <- var_df %>%
  gather(var, y)
ggplot(var_df_plot, aes(x = factor(var), y = y)) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(height = 0, 
    width = 0.25), fill = "lightseagreen", colour = "lightseagreen", alpha = 0.75, size = 3)
```

```{r}
# filter to just months 5 and 9
air5and9 <- airquality %>%
  filter(Month %in% c(5, 9))

air5and9 %>%
  group_by(Month) %>%
  dplyr::summarise(vars = var(Ozone, na.rm = TRUE),
            sds = sd(Ozone, na.rm = TRUE))
```

```{r}
# library(car)
# Levene's test
with(air5and9, leveneTest(Ozone ~ factor(Month), center = mean))
```

```{r}
# library(car)
# Brown-Forsythe
with(air5and9, leveneTest(Ozone ~ factor(Month), center = median))
```

```{r}
# filter to just months 8 and 9
air8and9 <- airquality %>%
  filter(Month %in% c(8, 9))

air8and9 %>%
  group_by(Month) %>%
  dplyr::summarise(vars = var(Ozone, na.rm = TRUE),
            sds = sd(Ozone, na.rm = TRUE))
```

```{r}
# library(car)
# Levene's test
with(air8and9, leveneTest(Ozone ~ factor(Month), center = mean))
```

```{r}
# library(car)
# Brown-Forsythe
with(air8and9, leveneTest(Ozone ~ factor(Month), center = median))
```

## Wilcoxon Mann Whitney (WMW) test for stochastic ordering of alternatives

```{r}
# library(datasets)
data("esoph")
head(esoph)
```

```{r}
# unit of analysis is records for 88 age/alcohol/tobacco combinations

# collapse across age/tobacco combinations
tidy_esoph <- esoph %>%
  group_by(alcgp) %>%
  dplyr::summarise(cases = sum(ncases),
            controls = sum(ncontrols)) %>%
  gather(group, n, -alcgp) %>%
  mutate(alcgp = as.numeric(alcgp)) # required for wilcoxon test

# need to create data with unit of analysis = participant
esoph_data <- tidy_esoph %>% 
  group_by(group) %>%
  do(data.frame(y = rep(.$alcgp, .$n)))

head(esoph_data)
```

```{r}
esoph_data %>%
  group_by(group, y) %>%
  tally()
```

```{r}
# run the test!
wilcox.test(y ~ group, data = esoph_data)
```

### Air quality data

```{r}
sum(1:55) #check with R
```

```{r}
min_rank_sum <- cbind(min_R_A=sum(1:26), min_R_B=sum(1:29))
min_rank_sum #This is what R subtracts from W
```

```{r}
pwilcox(377, 26, 29) #pi=P(W0) based on discrete distribution
```

```{r}
qwilcox(.5, 26, 29) #this would gotten us to W0 also
```

```{r}
qwilcox(.025, 26, 29)
```

```{r}
qwilcox(.975, 26, 29)
```

```{r}
air_ranks <- airquality %>%
  filter(Month %in% c(5, 9), !is.na(Ozone)) %>%
  mutate(oz_rank = rank(Ozone, ties.method = "average"))
# check that it worked
air_ranks %>%
  arrange(oz_rank) %>%
  head()
```

```{r}
obs_rank_sum <- air_ranks %>%
  group_by(Month) %>%
  dplyr::summarise(rank_sum = sum(oz_rank))
obs_rank_sum
```

```{r}
w_a <- obs_rank_sum$rank_sum[1] - min_rank_sum[1]
w_b <- obs_rank_sum$rank_sum[2] - min_rank_sum[2]
w_min <- min(w_a, w_b) #take the minimum as the test statistic
cbind(w_a, w_b, w_min)
```

```{r}
p_min <- min(pwilcox(w_min, 26, 29), 1-pwilcox(w_min, 26, 29))
p_2tailed <- 2*p_min
c(w_min, p_2tailed)
```

```{r}
w1 <- wilcox.test(Ozone ~ Month, data = airquality,
            subset = Month %in% c(5, 9), correct = FALSE, exact = TRUE)
```

```{r}
w1
```

```{r}
lowerz <- -1.559383
upperz <- -1.576241
pmin <- min(pnorm(lowerz), 1 - pnorm(upperz))
2*pmin
```

```{r}
# library(coin)
# coin is especially fussy about predictors as factors
air_ranks$Month <- as.factor(air_ranks$Month)
w2 <- wilcox_test(Ozone ~ Month, data = air_ranks, conf.int = TRUE, distribution = "exact")
w2
```

```{r}
# and right back where we started from!
w2@statistic@linearstatistic # the uncorrected observed rank sum for group A
```

```{r}
w2@statistic@linearstatistic - min_rank_sum[1] # the corrected observed W for group A
```

```{r}
# we can also check what W was under the null
expectation(w2) # should give you 728
```

```{r}
w_asymp <- wilcox_test(Ozone ~ Month, data = air_ranks, conf.int = TRUE)
w_asymp
```

```{r}
x <- air_ranks$Ozone[air_ranks$Month == 5] 
y <- air_ranks$Ozone[air_ranks$Month == 9] 
diffs <- sort(as.vector(outer(y, x, "-")))
median(diffs)
```

```{r}
# library(pairwiseCI)
# pairwiseCI can be especially fussy about predictors as factors
air_ranks$Month <- as.factor(air_ranks$Month)
pairwiseCI(Ozone ~ Month, data= air_ranks, method = "HL.diff") #Exact conditional nonparametric CI for difference of locations
```

```{r}
air_ranks %>%
  group_by(Month) %>%
  summarise(medians = median(Ozone))
```

```{r}
n_a <- 26
n_b <- 29
u_a <- n_a*n_b + min_rank_sum[1] - obs_rank_sum$rank_sum[1]
u_b <- n_a*n_b + min_rank_sum[2] - obs_rank_sum$rank_sum[2]
cbind(u_a, u_b, min(u_a, u_b))
```
  
## Your turn

The hypothesis that babies born to mothers who smoked have different birthweights than babies whose mothers did not smoke (2-tailed test).

```{r}
library(MASS)
data(birthwt)
```

```{r}
select <- dplyr::select
```


1. The total rank sum

```{r}
rank_sum <- sum(1:nrow(birthwt))
rank_sum
```

2. Expected rank sums for groups A and B under the null hypothesis

```{r}
na <- birthwt %>% filter(smoke==0) %>% nrow()
nb <- birthwt %>% filter(smoke==1) %>% nrow()

exp_R_A <- (na * (nrow(birthwt) + 1))/2
exp_R_B <- (nb * (nrow(birthwt) + 1))/2
exp_R_A
exp_R_B
```

3. Minimum rank sums for groups A and B

```{r}
min_rank_sum <- cbind(min_R_A=sum(1:na),min_R_B=sum(1:nb))
min_rank_sum
```

4. The expected (corrected) W statistic under the null hypothesis for groups A and B

```{r}
w_a_null <- exp_R_A - min_rank_sum[1] #qwilcox(.5,na,nb)
w_b_null <- exp_R_B - min_rank_sum[2] #qwilcox(.5,nb,na)
w_a_null
w_b_null
pwilcox(w_a,na,nb)
qwilcox(.5,na,nb)
```

5. The regions of rejection (α=.05, two tailed) for the expected (corrected) W statistic for groups A and B

```{r}
w_lo <- qwilcox(.025,na,nb)
w_hi <- qwilcox(.975,na,nb)
w_lo
w_hi
```

6. Observed (uncorrected) rank sums for groups A and B

```{r}
birthwt_ranks <- birthwt %>%
  filter(smoke %in% c(0,1), !is.na(bwt)) %>%
  mutate(bwt_rank = rank(bwt, ties.method = "average"))

birthwt_ranks %>% 
  arrange(bwt_rank) %>%
  head()

obs_rank_sum <- birthwt_ranks %>%
  group_by(smoke) %>%
  dplyr::summarise(rank_sum = sum(bwt_rank))

obs_rank_sum
```

7. Observed (corrected) rank sums for groups A and B

```{r}
w_a_obs <- obs_rank_sum$rank_sum[1] - min_rank_sum[1]
w_b_obs <- obs_rank_sum$rank_sum[2] - min_rank_sum[2]
w_a_obs
w_b_obs
```

8. The observed (minimum corrected) W statistic and its p-value

```{r}
w_min <- min(w_a_obs, w_b_obs) #take the minimum as the test statistic
p_min <- min(pwilcox(w_min, na, nb), 1-pwilcox(w_min, na, nb))
p_2tailed <- 2*p_min
c(w_min, p_2tailed)
```

9. The z-statistic and its p-value (either exact or asymptotic depending on your sample size)

```{r}
mu_w_a <- (na * (nrow(birthwt) + 1))/2
mu_w_a
sd_w_a <- (na * nb * (nrow(birthwt) + 1))/2
sd_w_a
z_a <- (w_a_obs + .5 - mu_w_a)/sqrt(sd_w_a)
z_a
lowerz <- z_a + .5
upperz <- z_a - .5
pmin <- min(pnorm(lowerz), 1 - pnorm(upperz))
2*pmin
```

10. The Hodges-Lehmann estimate and its 95% confidence interval

```{r}
birthwt_ranks$smoke <- as.factor(birthwt_ranks$smoke)
pairwiseCI(bwt ~ smoke, data= birthwt_ranks, method = "HL.diff")
```

11. What do the results suggest about your null hypothesis?

> The results suggest mothers who smoke during pregnancy give birth to children with lower median weights than mothers who don't smoke during pregnancy.
