---
title: "Math 530/630 Final Replication Project"
author: "Josh Burkhart, Steve Chamberlin, Kristen Stevens"
date: "December 2, 2015"
output: pdf_document
---

# Association of Arsenic Exposure with Lung Cancer Incidence Rates in the United States

Citation: Putila JJ, Guo NL (2001) Association of Arsenic Exposure with Lung Cancer Incidence Rates in the United States. PLOS ONE 6(10): e25886.

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
knitr::opts_chunk$set(error = TRUE, comment = NA, warnings = FALSE, messages = FALSE, tidy = TRUE, fig.path="Figs/", echo = TRUE, include = TRUE, digits = 3)
options(scipen=999)
```

```{r load_packages, include = FALSE}
suppressWarnings(suppressMessages(library(ggplot2)))
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(tidyr)))
```

```{r import_data}
full <- read.csv("./File_S4.csv", 
                 row.names = 1) # change directory for FileS4.csv
```

The authors' data file (FileS4.csv) includes 757 observations of 20 variables.  We have named this data.frame "full". The unit of analysis is County. The variables of interest are mean aresenic level in parts per million weighted by county population (Ascounty), median income of county (MedIncome), population (Population), county smoking prevalence calculated as a percent of respondents age 18 or older who reported having smoked more than 100 cigarettes in their lifetime (smkrate), and age-adjusted lung cancer incidence rates (AdjRate).

We first generated log transformed and centered variables from the variables described above. 

```{r log_transformed_centered_variables}
full$lnAs <- log(full$Ascounty) - mean(na.omit(log(full$Ascounty))) 
full$lnInc <- log(full$MedIncome) - mean(na.omit(log(full$MedIncome))) 
full$Population <- as.numeric(as.character(full$Population))
full$lnsmk <- full$smkrate
full$lnar <- log(full$AdjRate)
```

## Regression Analysis

"The first analysis sought to determine the influence of exposure levels of arsenic on lung cancer incidence in the U.S., and persistance of these effects controlling for possible confounders. The association betweeen each contaminant and lung cancer incidence was assessed using Poisson regression in order to reflect the annual incidence rate as a counting measure."

For comparison, we performed a linear regression in addition to a Poisson regression for each analysis. The following models predict lung cancer incidence weighted by county population using the untransformed data.

```{r unadjusted_model}
# Poisson regression
glm1 <- glm(full$AdjRate ~ full$Ascounty, 
            family=poisson, 
            weights=as.numeric(full$Population)) 
summary(glm1)
# odds ratio with CI
glm1_odds <- exp(cbind(OR = coef(glm1), confint(glm1))) 

# Linear regression
lm1 <- lm(full$lnar ~ full$Ascounty,
          weights=as.numeric(full$Population))
summary(lm1)
# odds ratio with CI
lm1_odds <- exp(cbind(OR = coef(lm1), confint(lm1))) 
```

```{r model_adjusted_for_smoking_and_income}
# Poisson regression
SESassmk <- glm(full$AdjRate ~ full$smkrate 
                + full$Ascounty 
                + full$MedIncome, 
                family=poisson, 
                weights=as.numeric(full$Population))
summary(SESassmk)
SESassmk_odds <- exp(cbind(OR = coef(SESassmk), confint(SESassmk)))

#Linear regressions
SESassmklm <- lm(full$lnar ~ full$smkrate 
                 + full$Ascounty 
                 + full$MedIncome,
                 weights=as.numeric(full$Population))
summary(SESassmklm)
SESassmklm_odds <- exp(cbind(OR = coef(SESassmklm), confint(SESassmklm)))

```

## Table 1: Unadjusted Model

Summary of Poisson regressions of the effect of arsenic concentration (ppm) on county-level lung cancer incidence rates in the U.S. in an unadjusted model. (We broke up Table 1 into two parts: the Unadjusted Model and the Adjusted Model.)

```{r table_1_unadjusted_model}
Model_and_Variable = c("Arsenic")
Coefficient = c(round(summary(glm1)$coefficients[2], digits = 4))
Std.Error = c("1.9 x 10^-6") # summary(glm1)$coefficients[4]
Odds_Ratio_CI = c(paste(c(round(glm1_odds[2,1], digits = 3), 
                          "(", round(glm1_odds[2,2], digits = 3), "-",
                          round(glm1_odds[2,3], digits = 3), ")") 
                        ,collapse = " "))
P_value = c("P<0.0001")
N = c(742)
df = data.frame(Model_and_Variable,Coefficient,Std.Error,Odds_Ratio_CI,P_value,N)
kable(df)
# should we add the P-value and N as variables?
# remove underscores from table labels?
# table in knit pdf needs to be reformatted to fit the page
# scientific notation?
# do we want to add a table for the results from the linear model?
```

## Table 1: Adjusted Model

Summary of Poisson regressions of the effect of arsenic concentration (ppm) on county-level lung cancer incidence rates in the U.S. in a model adjusted for both smoking and median county income.

```{r table_1_adjusted}
Model_and_Variable = c("Arsenic","Smoking","Median Income")
Coefficient = c(round(summary(SESassmk)$coefficients[3],digits = 4),
                round(summary(SESassmk)$coefficients[2], digits = 4),
                round(summary(SESassmk)$coefficients[4], digits = 4))
Std.Error = c("1.9 x 10^-6",
              "0.0002",
              "1.31 x 10^-9")
Odds_Ratio_CI = c(paste(c(round(SESassmk_odds[3], digits = 3), "(", 
                            round(SESassmk_odds[7], digits = 3), "-", 
                            round(SESassmk_odds[11], digits = 3), ")") 
                          ,collapse = " "),
                  paste(c(round(SESassmk_odds[2], digits = 3), "(", 
                            round(SESassmk_odds[6], digits = 3), "-", 
                            round(SESassmk_odds[10], digits = 3), ")") 
                          ,collapse = " "),
                  paste(c(round(SESassmk_odds[4], digits = 3), "(", 
                            round(SESassmk_odds[8], digits = 3), "-", 
                            round(SESassmk_odds[12], digits = 3), ")") 
                          ,collapse = " "))
P_value = c("P<0.0001")
N = c(585)
df = data.frame(Model_and_Variable,Coefficient,Std.Error,Odds_Ratio_CI,P_value,N)
kable(df)
# should we add the P-value and N as variables?
# remove underscores from table labels?
# table in knit pdf needs to be reformatted to fit the page
# scientific notation?
# do we want to add a table for the results from the linear model?
```

## Table 2

Difference in lung cancer incidence attributable to arsenic exposure alone for high and low-exposure areas in the U.S. based on the results of the adjusted Poisson models and the USGS survey quantiles in Figure 1. 

```{r table_2}
Compound = c("Arsenic")
Low_ppm = c(1.477)
High_ppm = c(14.525)
B_Estimate = c(0.0039)
Lung_Cancer_Rate_Increase_Pct = c("5.3%")
df = data.frame(Compound,Low_ppm,High_ppm,B_Estimate,Lung_Cancer_Rate_Increase_Pct)
kable(df)
# change values in table to variables, need to move this.
```

## Figure 2

**Description, explanation of variables, and code clean-up needed here.

```{r }
## Estimate the 25, 50, amd 75% quartile points for each variable for the quartiles interaction models
AsCut <- NA
AsCut[1] <- as.numeric(summary(full$lnAs)[2])
AsCut[2] <- as.numeric(summary(full$lnAs)[3])
AsCut[3] <- as.numeric(summary(full$lnAs)[5])

SmkCut <- NA
SmkCut[1] <- as.numeric(summary(full$lnsmk)[2])
SmkCut[2] <- as.numeric(summary(full$lnsmk)[3])
SmkCut[3] <- as.numeric(summary(full$lnsmk)[5])

SESCut <- NA
SESCut[1] <- as.numeric(summary(full$lnInc)[2])
SESCut[2] <- as.numeric(summary(full$lnInc)[3])
SESCut[3] <- as.numeric(summary(full$lnInc)[5])

## Calculate Strat Groups ##
## Smoking Quartiles
smkgrp <- ifelse(is.na(full$lnsmk), NA, ifelse(full$lnsmk < SmkCut[1], 1, ifelse(full$lnsmk >= SmkCut[1] & full$lnsmk < SmkCut[2], 2, ifelse(full$lnsmk >= SmkCut[2] & full$lnsmk < SmkCut[3], 3, 4))))
## SES Low-Income Cutoffs
SESgrp <- ifelse(is.na(full$MedIncome), NA, ifelse(full$MedIncome < 24000 & !is.na(full$MedIncome), 1, ifelse(full$MedIncome>= 24000 & full$MedIncome< 28700, 2, ifelse(full$MedIncome >= 28700 & full$MedIncome < 38300, 3, 4))))
## SES Quartiles
#SESgrp <- ifelse(full$lnInc< -0.158, 1, ifelse(full$lnInc>=-0.158 & full$lnInc< -0.00391, 2, ifelse(full$lnInc>=-0.00391 & full$lnInc <0.1478, 3, 4)))
## Arsenic Quartiles
AsQ <- ifelse(is.na(full$lnAs), NA, ifelse(full$lnAs < AsCut[1], 1, ifelse(full$lnAs >= AsCut[1] & full$lnAs < AsCut[2], 2, ifelse(full$lnAs >= AsCut[2] & full$lnAs < AsCut[3], 3, 4))))

full$smkgrp <- ifelse(is.na(full$lnsmk), NA, ifelse(full$lnsmk < SmkCut[1], 1, ifelse(full$lnsmk >= SmkCut[1] & full$lnsmk < SmkCut[2], 2, ifelse(full$lnsmk >= SmkCut[2] & full$lnsmk < SmkCut[3], 3, 4))))
## SES Low-Income Cutoffs
full$SESgrp <- ifelse(is.na(full$MedIncome), NA, ifelse(full$MedIncome < 24000 & !is.na(full$MedIncome), 1, ifelse(full$MedIncome>= 24000 & full$MedIncome< 28700, 2, ifelse(full$MedIncome >= 28700 & full$MedIncome < 38300, 3, 4))))
## SES Quartiles
#SESgrp <- ifelse(full$lnInc< -0.158, 1, ifelse(full$lnInc>=-0.158 & full$lnInc< -0.00391, 2, ifelse(full$lnInc>=-0.00391 & full$lnInc <0.1478, 3, 4)))
## Arsenic Quartiles
full$AsQ <- ifelse(is.na(full$lnAs), NA, ifelse(full$lnAs < AsCut[1], 1, ifelse(full$lnAs >= AsCut[1] & full$lnAs < AsCut[2], 2, ifelse(full$lnAs >= AsCut[2] & full$lnAs < AsCut[3], 3, 4))))

## Quartile-Based Interaction Models
## Convert quartiles to factors
AsQf <- as.factor(AsQ)
smkgrpf <- as.factor(smkgrp)
smkgrpfbak <- smkgrpf
SESgrpf <- as.factor(SESgrp)

full$AsQf <- as.factor(full$AsQ)
full$smkgrpf <- as.factor(full$smkgrp)
full$smkgrpfbak <- full$smkgrpf
full$SESgrpf <- as.factor(full$SESgrp)
```

```{r }
#############
## ARSENIC ## figure 2
#############
## Arsenic and Smoking This creates the data for the first line of table 3 and the p
## value displayed on the right graph on figure 2
smkgrpf <- smkgrpfbak
smkgrpf <- ifelse(is.na(smkgrpf), NA, ifelse(smkgrpf==1 | smkgrpf==2, 1, 2))
intAsSmk <- aov(full$AdjRate ~ SESgrpf + AsQf*smkgrpf, weights=as.numeric(full$Population))
summary(intAsSmk)

## Without SES This creates the p value displayed on the left graph for figure 2
intAsSmk2 <- aov(full$AdjRate ~ AsQf*smkgrpf, weights=as.numeric(full$Population))
summary(intAsSmk2)

## Arsenic and SES this creates the statistics for table3 second model
intAsSES <- aov(full$AdjRate ~ smkgrpf + AsQf*SESgrpf, weights=as.numeric(full$Population))
summary(intAsSES) 
```

```{r}
## Plot the Interaction between Arsenic and Smoking WITHOUT SES
smkgrpf <- smkgrpfbak
smkgrpf <- ifelse(is.na(smkgrpf), NA, ifelse(smkgrpf==1 | smkgrpf==2,1,2))

# Two models are run to get the two lines for low and high smoking, unadjusted
r1 <- glm(full[smkgrpf==1,]$AdjRate ~ full[smkgrpf==1,]$lnAs, family=poisson, weights=as.numeric(full[smkgrpf==1,]$Population))
summary(r1)
r2 <- glm(full[smkgrpf==2,]$AdjRate ~ full[smkgrpf==2,]$lnAs, family=poisson, weights=as.numeric(full[smkgrpf==2,]$Population)) # run model to get the fitted values
summary(r2)

	#	this section builds the dataset from the model output for graphing
data1 <- cbind( c(t(r1$model[2]),t(r2$model[2])), # ln arsenic values
			        	c(log(r1$fitted.values),log(r2$fitted.values)), # fitted dependent vars
			        	c(r1$weights, r2$weights), # county populations
			        	c(rep("1",dim(r1$model[2])[1]), rep("2",dim(r2$model[2])[1]))) 
			        	# smoking group

data1 <- as.data.frame(data1, stringsAsFactors=FALSE)
names(data1) <- c("logAs","logRate", "weight", "smkgrp")
data1$logAs  <- as.numeric(as.character(data1$logAs))
data1$logRate <- as.numeric(as.character(data1$logRate)) # these are the fitted values
data1$weight  <- as.numeric(as.character(data1$weight))
data1$smkgrp  <- as.numeric(as.character(data1$smkgrp))
data1$adjinc  <- c(as.numeric(coef(r1)[2])*r1$model[,2]*0, as.numeric(coef(r2)[2])*r2$model[,2]*0)  # this is the increment added by arsenic, but it
                                         # is zeroed out
data1$adjrate <- data1$adjinc+data1$logRate # the same as logRate

#This creates the graph but I'm not sure if you can have the fitted line without dots

assmkp <- ggplot(data1, aes(x=logAs, y=adjrate, shape=factor(smkgrp), color=factor(smkgrp))) 
assmkp + stat_smooth(method = "glm", level=0.95, alpha=1, fill="grey80", color="black") +
		 #scale_color_manual(values=c("grey50","grey70")) +		
		 geom_point(aes(size=weight)) +
		 geom_point() +
		 theme(legend.position = "right") +
		 theme_bw()

full$smkgrpf <- full$smkgrpfbak
full$smkgrpf <- ifelse(is.na(full$smkgrpf), NA, ifelse(full$smkgrpf==1 | full$smkgrpf==2, 1, 2))

full$SESgrp2f <- full$SESgrpf
full$SESgrp2f <- ifelse(is.na(full$SESgrp2f), NA, ifelse(full$SESgrp2f==1 | full$SESgrp2f==2, 1, 2))

# This creates the combined SES and smoking group into one variable, categories 1,2 
# are the low smoking categories, important to note that the SES categories are not
# quartiles

full$sessmk = ifelse(full$SESgrp2f==1 & full$smkgrpf==1,"1) Low SES, Low Smoke",
              ifelse(full$SESgrp2f==2 & full$smkgrpf==1,"2) High SES, Low Smoke", 
              ifelse(full$SESgrp2f==1 & full$smkgrpf==2,"3) Low SES, High Smoke", 
            ifelse(full$SESgrp2f==2 & full$smkgrpf==2,"4) High SES, High Smoke",NA))))
 
# This is the graph with only two lines for the two smoking groups with actual data

full$smklabel <- ifelse(is.na(full$smkgrpf), NA, ifelse(full$smkgrpf==1,'1) Low Smoke', '2) High Smoke'))

graphfile <- full %>%
             group_by(AsQf,smklabel) %>%
             summarise(meanrate = median(AdjRate) ) 

finalgraph <- graphfile[!is.na(graphfile$AsQf) & !is.na(graphfile$smklabel),]

ggplot(finalgraph, aes(x=AsQf, y=meanrate, color=factor(smklabel))) +
         geom_line(aes(group=smklabel)) +
    ggtitle("Lung Cancer Incidence by Arsenic Quartiles, by Smoking Cat") +
    xlab("Arsenic Quartiles") +
    ylab("Lung Cancer Incidence (per 100,000")

# This graph creates four lines for the combination of SES and smoking

graphfile <- full %>%
             group_by(AsQf,sessmk) %>%
             summarise(meanrate = median(AdjRate) ) 

finalgraph <- graphfile[!is.na(graphfile$AsQf) & !is.na(graphfile$sessmk),]

ggplot(finalgraph, aes(x=AsQf, y=meanrate, color=factor(sessmk))) +
         geom_line(aes(group=factor(sessmk))) +
    ggtitle("Lung Cancer Incidence by Arsenic Qs, by Smoking/SES Cat") +
    xlab("Arsenic Quartiles") +
    ylab("Lung Cancer Incidence (per 100,000")    
  
```


```{r}
## GLMS for smoking levels WITH SES
smkgrpf <- smkgrpfbak
## Bottom 50% vs Top 50%
smkgrpf <- ifelse(is.na(smkgrpf), NA, ifelse(smkgrpf==1 | smkgrpf==2, 1, 2))

r1 <- glm(full[smkgrpf==1,]$AdjRate ~ full[smkgrpf==1,]$lnAs + full[smkgrpf==1,]$MedIncome, family=poisson, weights=as.numeric(full[smkgrpf==1,]$Population))
summary(r1)
r2 <- glm(full[smkgrpf==2,]$AdjRate ~ full[smkgrpf==2,]$lnAs + full[smkgrpf==2,]$MedIncome, family=poisson, weights=as.numeric(full[smkgrpf==2,]$Population))
summary(r2)

## Plot the Interaction between Arsenic and Smoking with SES
data1 <- cbind( c(t(r1$model[2]),t(r2$model[2])), # ln arsenic values
				c(log(r1$fitted.values),log(r2$fitted.values)), # fitted dependent variables
				c(r1$weights, r2$weights),  # county populations
				c(rep("1",dim(r1$model[2])[1]), rep("2",dim(r2$model[2])[1]))) # number of                                                                # counties in each smoking group
data1 <- as.data.frame(data1, stringsAsFactors=FALSE)
names(data1) <- c("logAs","logRate", "weight", "smkgrp")
data1$logAs  <- as.numeric(as.character(data1$logAs))
data1$logRate <- as.numeric(as.character(data1$logRate))
data1$weight  <- as.numeric(as.character(data1$weight))
data1$smkgrp  <- as.numeric(as.character(data1$smkgrp))
data1$adjinc  <- c(as.numeric(coef(r1)[2])*r1$model[,2], as.numeric(coef(r2)[2])*r2$model[,2]) # this is the increment added by arsenic
data1$adjrate <- data1$adjinc+data1$logRate
#data1$adjrate <- data1$logRate # turn off the adjustment to the fitted values

assmkp <- ggplot(data1, aes(x=logAs, y=adjrate, shape=factor(smkgrp), color=factor(smkgrp))) 
assmkp + stat_smooth(method = "glm", level=0.95, alpha=1, fill="grey80", color="black") +
		 #scale_color_manual(values=c("grey50","grey70")) +		
		 geom_point(aes(size=weight)) +
  	 geom_point() +
		 theme(legend.position = "right") +
		theme_bw()

```

## Table 3

Summary of ANOVA tests performed between Arsenic and covariates used in the regression analysis.

```{r table_3}
Interaction_Pair = c("Arsenic:Smoking","Arsenic:MCI")
DF = c(3,9)
F_value = c(2.6595,1.7798)
P_value = c(0.04747, 0.06914)
df = data.frame(Interaction_Pair, DF, F_value, P_value)
kable(df)
# change values in table to variables
```

## Figure 3

Combination violin and boxplots showing the average level of exposure and outcomes for counties in West Virginia or Kentucky comared with the remaining 10 states in the original sample.

Instead of the bar plot, we represented the data in Figure 3 with a series of combination violin and boxplots. The authors of the original paper attempted to represent each of the variables of interest for this data on the same y-axis scale, including arsenic, income, lung cancer incidence rates, and smoking.  Although the authors likely chose this visualization to save space, we decided that it was somewhat confusing and that additional information about the data distributions could be added if we instead used a combination of violin and boxplots.

**Explanation of variables here?
**Add x label (location) and y label (arsenic, income, etc.)

```{r arsenic_violin_plot}
# arsenic
KYWVas <- na.omit(full[full$SFIPS==21 | full$SFIPS==54,]$Ascounty)
notKYWVas <- na.omit(full[full$SFIPS!=21 & full$SFIPS!=54,]$Ascounty)
KYWVdf <- KYWVas %>% 
  as.data.frame() %>%
  mutate(KYWV="WV/KY") %>% 
  select("Ascounty"=1,"KYWV"=2)
notKYWVdf <- notKYWVas %>% 
  as.data.frame() %>%
  mutate(KYWV="Other") %>% 
  select("Ascounty"=1,"KYWV"=2)
KYWVcombined = merge(KYWVdf, notKYWVdf, all=TRUE)
KYWVcombined %>% 
  ggplot(aes(x=KYWV,y=Ascounty)) +
  geom_violin(aes(fill = factor(KYWV))) +
  geom_boxplot(width=0.1)
```

```{r smoking_violin_plot}
# smoking
KYWVsmk <- na.omit(full[full$SFIPS==21 | full$SFIPS==54,]$smkrate)
notKYWVsmk <- na.omit(full[full$SFIPS!=21 & full$SFIPS!=54,]$smkrate)
KYWVdf <- KYWVsmk %>% 
  as.data.frame() %>%
  mutate(KYWV="WV/KY") %>% 
  select("smkrate"=1,"KYWV"=2)
notKYWVdf <- notKYWVsmk %>% 
  as.data.frame() %>%
  mutate(KYWV="Other") %>% 
  select("smkrate"=1,"KYWV"=2)
KYWVcombined = merge(KYWVdf,notKYWVdf,all=TRUE)
KYWVcombined %>% 
  ggplot(aes(x=KYWV,y=smkrate)) +
  geom_violin(aes(fill = factor(KYWV))) +
  geom_boxplot(width=0.1)
```

```{r income_violin_plot}
# income
KYWVmed <- na.omit(full[full$SFIPS==21 | full$SFIPS==54,]$MedIncome)
notKYWVmed <- na.omit(full[full$SFIPS!=21 & full$SFIPS!=54,]$MedIncome) 
KYWVdf <- KYWVmed %>% 
  as.data.frame() %>%
  mutate(KYWV="WV/KY") %>% 
  select("MedIncome"=1,"KYWV"=2)
notKYWVdf <- notKYWVmed %>% 
  as.data.frame() %>%
  mutate(KYWV="Other") %>% 
  select("MedIncome"=1,"KYWV"=2)
KYWVcombined = merge(KYWVdf,notKYWVdf,all=TRUE)
KYWVcombined %>% 
  ggplot(aes(x=KYWV,y=MedIncome)) +
  geom_violin(aes(fill = factor(KYWV))) +
  geom_boxplot(width=0.1)
```

```{r lung_cancer_incidence_rate_violin_plot}
# lung cancer incidence rate
KYWVrate <- na.omit(full[full$SFIPS==21 | full$SFIPS==54,]$AdjRate)
notKYWVrate	<- na.omit(full[full$SFIPS!=21 & full$SFIPS!=54,]$AdjRate) 
KYWVdf <- KYWVrate %>% 
  as.data.frame() %>%
  mutate(KYWV="WV/KY") %>% 
  select("AdjRate"=1,"KYWV"=2)
notKYWVdf <- notKYWVrate %>% 
  as.data.frame() %>%
  mutate(KYWV="Other") %>% 
  select("AdjRate"=1,"KYWV"=2)
KYWVcombined = merge(KYWVdf,notKYWVdf,all=TRUE)
KYWVcombined %>% 
  ggplot(aes(x=KYWV,y=AdjRate)) + 
  geom_violin(aes(fill = factor(KYWV))) +
  geom_boxplot(width=0.1)
```

## Extension

**Add extension here.