---
title: "hw5"
author: "Joshua Burkhart"
date: "December 8, 2015"
output: 
  pdf_document: 
    latex_engine: xelatex
---

# Homework 5: ANOVA

```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)
library(car)
library(datasets)
library(coin)
library(pairwiseCI)
library(broom)
library(phia)
library(Rfit)
options(scipen=999)
```

## Analysis of the expression of an isoform of the electroneutral Na+/H+ exchanger (NHE3)

Re-create the following two plots (note: don't worry about colors, but you should be able to make plots that have the same x/y axes, facetting, and geoms using ggplot2):

```{r 2}
# for comma separated values
stickle <- read.table("~/Software/Probability/hw5/qPCR_data.csv",
                        header = TRUE, sep = ",")
stickle$Temperature <- as.factor(stickle$Temperature)
```

```{r}
stickle %>%
  group_by(Temperature) %>%
  ggplot() +
  facet_wrap(~ Ecotype) +
  geom_boxplot(aes(x=Temperature, y=NHE3,color=Temperature)) +
  geom_point(aes(x=Temperature, y=NHE3,color=Temperature)) +
  scale_color_brewer(palette="Paired") + 
  xlab("Temperature") +
  theme(legend.position="none")
```

```{r}
stickle %>%
  group_by(Temperature) %>%
  ggplot() +
  facet_wrap(~ Ecotype) +
  geom_violin(aes(x=Temperature, y=NHE3,color=Temperature)) +
  geom_point(aes(x=Temperature, y=NHE3,color=Temperature)) +
  scale_color_brewer(palette="Paired") +
  xlab("Temperature") +
  theme(legend.position="none")
```

## On your own

This summary of results appears to focus on a two-way ANOVA predicting NHE3 gene expression levels from Ecotype and Temperature. Run this model on your own: interpret the regression coefficients from the general linear model, using contr.sum for all factor predictors, and conduct any post-hoc comparisons as you see appropriate to aid in interpreting the model.

```{r}
nhe3_sum <- lm(NHE3 ~ Temperature * Ecotype, 
               data = stickle, 
               contrasts = list(Ecotype = contr.sum,
                                Temperature = contr.sum))
tidy(nhe3_sum)
attr(model.matrix(nhe3_sum), "contrasts") # contrast check

#interpreting regression coefficients
grandmean <- mean(stickle$NHE3)

# main effects
bTemp1 <- stickle %>% 
  filter(Temperature == 4) %>% 
  summarise(bTemp1 = mean(NHE3) - grandmean)
Ecotype1 <- stickle %>% 
  filter(Ecotype == "Hybrid") %>% 
  summarise(bEcotype1 = mean(NHE3) - grandmean)
Ecotype2 <- stickle %>% 
  filter(Ecotype == "Lake") %>% 
  summarise(bEcotype2 = mean(NHE3) - grandmean)

cbind(grandmean, bTemp1, Ecotype1, Ecotype2)

Anova(nhe3_sum,type=2,white.adjust=TRUE)
```

> It appears the mean NHE3 differed depending on Temperature but
  not Ecotype.
  
```{r}
nhe3_twowayvars <- augment(nhe3_sum)
nhe3_resid <- ggplot(data = nhe3_twowayvars, aes(x = .fitted, 
                                                  y = .resid, 
                                                  colour = Ecotype)) +
  geom_point(aes(shape = Temperature), size = 5, position = position_jitter(width = .02)) +
  scale_colour_brewer(type = "qual", palette = 3) +
  theme_bw() 
suppressWarnings(print(nhe3_resid))
```

```{r}
plot(nhe3_sum)
```

```{r}
testInteractions(nhe3_sum, pairwise = "Temperature", adjustment = "BH")
testInteractions(nhe3_sum, pairwise = "Temperature", adjustment = "holm")
plot(interactionMeans(nhe3_sum))
testInteractions(nhe3_sum, fixed = "Ecotype", pairwise = "Temperature", adjustment = "bonferroni")
```

> The authors' statement, "substantial change with temperature in the
  marine ecotype that was not observed in the freshwater ecotype under
  the conditions tested here," seems to be supported by our interaction
  test. Indeed, the freshwater ecotype does not exhibit a significant
  interaction between temperature and NHE3 expression, while the marine 
  ecotype does.











